Differential abundance analysis of taxonomically-biased microbiome measurements

### 2022-03-13 abundance-measurement.Rmd
Change equation label

Our model describes the mathematical relationship between the read counts obtained by MGS and the true absolute abundances of each species in a sample.
For concreteness, we take _absolute abundance_, or simply _abundance_, to refer to cell concentration, or number of cells per unit volume, in a sample.
Thus we take 'absolute' simply to mean not relative to other species; however, our results also apply were 'absolute' taken to mean all cells in a sample or in a sampled ecosystem.
Similarly, our results apply to abundance units other than cells, such as biomass and genome copy number.
Our model stipulates that the assigned read count of a species $i$ in a sample $a$ equals its abundance multiplied by species-specific and sample-specific factors,
\begin{align}
  (\#eq:measurement-model)
  \text{reads}_{i}(a)
  = \text{abun}_{i}(a) \quad \cdot
    \underbrace{\text{efficiency}_{i}}_{\substack{\text{species specific,} \\  \text{sample independent}}}
    \cdot \quad
    \underbrace{\text{effort}(a)}_{\substack{\text{species independent,} \\  \text{sample specific}}}.
\end{align}
The species-specific factor, $\text{efficiency}_{i}$, equals the *relative measurement efficiency* (or simply *efficiency*) of the species---how much more easily that species is measured (converted from cells to assigned reads) relative to a arbitrary fixed reference species (@mclaren2019cons).
We assume the efficiencies of particular species are consistent across samples.
The variation in efficiency among species corresponds to the taxonomic bias of the MGS protocol.
The sample-specific factor, $\text{effort}(a)$, quantifies the *effective experimental effort* per unit abundance in measuring the sample.
<!-- It accounts for variation in sample input volume and total sequencing run output, as well as the mean species efficiency within the sample. -->
Equation \@ref(eq:measurement-model) implies that the total number of assigned reads in sample $a$ equals
\begin{align}
  (\#eq:total-reads)
  \text{reads}_S(a)
    = \text{abun}_S(a) \cdot \text{efficiency}_S(a) \cdot \text{effort}(a),
\end{align}
where $\text{abun}_{S}(a) \equiv \sum_{j\in S}\text{abun}_j(a)$ is the total abundance of species in $S$ and 
\begin{align}
  (\#eq:mean-efficiency)
  \text{efficiency}_S(a) 
    \equiv \frac{\sum_{j\in S}\text{abun}_j(a) \cdot \text{efficiency}_j}{\text{abun}_S(a)}
\end{align}
is the _mean efficiency_ over all species in $S$ weighted by their abundance.
<!-- The subscript $S$ serves as a reminder that many reads may not be assigned---for example, reads that fail to pass quality control or that come from the host or other taxonomic domains than what is being studied. -->

@@

The proportion of a species is typically measured by its proportion of assigned reads,
\begin{align}
  (\#eq:prop-meas)
  \widehat{\text{prop}}_{i}(a) = \frac{\text{reads}_i(a)}{\text{reads}_S(a)}.
\end{align}
We can rewrite the right-hand side (using Equations \@ref(eq:measurement-model), \@ref(eq:total-reads), and \@ref(eq:prop-meas)) to find
\begin{align}
  (\#eq:prop-error)
  \widehat{\text{prop}}_{i}(a)
  &= \text{prop}_{i}(a) \cdot \underbrace{\frac{\text{efficiency}_{i}}{\text{efficiency}_S(a)}}_{\substack{\text{variable} \\ \text{fold error}}}.
\end{align}
Taxonomic bias creates a multiplicative error in the species' proportion equal to its efficiency relative to the mean efficiency in the sample.
Consequently, the species's proportion is measured as too high in samples that are dominated by species with lower efficiencies, and measured as too low in samples that are dominated by species with higher efficiencies.
This phenomenon is illustrated in two hypothetical communities in Figure \@ref(fig:error-proportions). 
Species 3 has an efficiency of 6; it is under-measured in Sample 1, which has a mean efficiency of 8.33, but over-measured in Sample 2, which has a mean efficiency of 3.15.
A demonstration in artificially constructed ('mock') communities of bacterial cells is shown in [Figure 3C](https://doi.org/10.7554/eLife.46923.004) of @mclaren2019cons.

The measured ratio between species $i$ and $j$ is given by the ratio of their read counts, 
\begin{align}
  (\#eq:ratio-meas)
  \widehat{\text{ratio}}_{i/j}(a) = \frac{\text{reads}_i(a)}{\text{reads}_j(a)}.
\end{align}
From Equations \@ref(eq:measurement-model) and \@ref(eq:ratio-meas), it follows that 
\begin{align}
  (\#eq:ratio-error)
  \widehat{\text{ratio}}_{i/j}(a)
%  &= \frac{\text{abun}_{i}(a)}{\text{abun}_{j}(a)} 
  &= \text{ratio}_{i/j}(a) \cdot \underbrace{\frac{\text{efficiency}_{i}}{\text{efficiency}_{j}}}_{\substack{\text{constant} \\ \text{fold error}}}.
\end{align}
Taxonomic bias creates a multiplicative measurement error in the ratio that is equal to the ratio in their efficiencies; the error is therefore constant across samples.
For instance, in Figure \@ref(fig:error-proportions), the ratio of Species 3 (with an efficiency of 6) to Species 1 (with an efficiency of 1) is over-estimated by a factor of 6 in both communities despite their varying compositions.
A demonstration in bacterial mock communities is shown in [Figure 3D](https://doi.org/10.7554/eLife.46923.004) of @mclaren2019cons.

>>>>

Our model describes the mathematical relationship between the read counts obtained by MGS and the true absolute abundances of each species in a sample.
For concreteness, we take _absolute abundance_, or simply _abundance_, to refer to cell concentration, or number of cells per unit volume, in a sample.
Thus we take 'absolute' simply to mean not relative to other species; however, our results also apply were 'absolute' taken to mean all cells in a sample or in a sampled ecosystem.
Similarly, our results apply to abundance units other than cells, such as biomass and genome copy number.
Our model stipulates that the assigned read count of a species $i$ in a sample $a$ equals its abundance multiplied by species-specific and sample-specific factors,
\begin{align}
  (\#eq:mgs-model)
  \text{reads}_{i}(a)
  = \text{abun}_{i}(a) \quad \cdot
    \underbrace{\text{efficiency}_{i}}_{\substack{\text{species specific,} \\  \text{sample independent}}}
    \cdot \quad
    \underbrace{\text{effort}(a)}_{\substack{\text{species independent,} \\  \text{sample specific}}}.
\end{align}
The species-specific factor, $\text{efficiency}_{i}$, equals the *relative measurement efficiency* (or simply *efficiency*) of the species---how much more easily that species is measured (converted from cells to assigned reads) relative to a arbitrary fixed reference species (@mclaren2019cons).
We assume the efficiencies of particular species are consistent across samples.
The variation in efficiency among species corresponds to the taxonomic bias of the MGS protocol.
The sample-specific factor, $\text{effort}(a)$, quantifies the *effective experimental effort* per unit abundance in measuring the sample.
<!-- It accounts for variation in sample input volume and total sequencing run output, as well as the mean species efficiency within the sample. -->
Equation \@ref(eq:mgs-model) implies that the total number of assigned reads in sample $a$ equals
\begin{align}
  (\#eq:total-reads)
  \text{reads}_S(a)
    = \text{abun}_S(a) \cdot \text{efficiency}_S(a) \cdot \text{effort}(a),
\end{align}
where $\text{abun}_{S}(a) \equiv \sum_{j\in S}\text{abun}_j(a)$ is the total abundance of species in $S$ and 
\begin{align}
  (\#eq:mean-efficiency)
  \text{efficiency}_S(a) 
    \equiv \frac{\sum_{j\in S}\text{abun}_j(a) \cdot \text{efficiency}_j}{\text{abun}_S(a)}
\end{align}
is the _mean efficiency_ over all species in $S$ weighted by their abundance.
<!-- The subscript $S$ serves as a reminder that many reads may not be assigned---for example, reads that fail to pass quality control or that come from the host or other taxonomic domains than what is being studied. -->

@@

The proportion of a species is typically measured by its proportion of assigned reads,
\begin{align}
  (\#eq:prop-meas)
  \widehat{\text{prop}}_{i}(a) = \frac{\text{reads}_i(a)}{\text{reads}_S(a)}.
\end{align}
We can rewrite the right-hand side (using Equations \@ref(eq:mgs-model), \@ref(eq:total-reads), and \@ref(eq:prop-meas)) to find
\begin{align}
  (\#eq:prop-error)
  \widehat{\text{prop}}_{i}(a)
  &= \text{prop}_{i}(a) \cdot \underbrace{\frac{\text{efficiency}_{i}}{\text{efficiency}_S(a)}}_{\substack{\text{variable} \\ \text{fold error}}}.
\end{align}
Taxonomic bias creates a multiplicative error in the species' proportion equal to its efficiency relative to the mean efficiency in the sample.
Consequently, the species's proportion is measured as too high in samples that are dominated by species with lower efficiencies, and measured as too low in samples that are dominated by species with higher efficiencies.
This phenomenon is illustrated in two hypothetical communities in Figure \@ref(fig:error-proportions). 
Species 3 has an efficiency of 6; it is under-measured in Sample 1, which has a mean efficiency of 8.33, but over-measured in Sample 2, which has a mean efficiency of 3.15.
A demonstration in artificially constructed ('mock') communities of bacterial cells is shown in [Figure 3C](https://doi.org/10.7554/eLife.46923.004) of @mclaren2019cons.

The measured ratio between species $i$ and $j$ is given by the ratio of their read counts, 
\begin{align}
  (\#eq:ratio-meas)
  \widehat{\text{ratio}}_{i/j}(a) = \frac{\text{reads}_i(a)}{\text{reads}_j(a)}.
\end{align}
From Equations \@ref(eq:mgs-model) and \@ref(eq:ratio-meas), it follows that 
\begin{align}
  (\#eq:ratio-error)
  \widehat{\text{ratio}}_{i/j}(a)
%  &= \frac{\text{abun}_{i}(a)}{\text{abun}_{j}(a)} 
  &= \text{ratio}_{i/j}(a) \cdot \underbrace{\frac{\text{efficiency}_{i}}{\text{efficiency}_{j}}}_{\substack{\text{constant} \\ \text{fold error}}}.
\end{align}
Taxonomic bias creates a multiplicative measurement error in the ratio that is equal to the ratio in their efficiencies; the error is therefore constant across samples.
For instance, in Figure \@ref(fig:error-proportions), the ratio of Species 3 (with an efficiency of 6) to Species 1 (with an efficiency of 1) is over-estimated by a factor of 6 in both communities despite their varying compositions.
A demonstration in bacterial mock communities is shown in [Figure 3D](https://doi.org/10.7554/eLife.46923.004) of @mclaren2019cons.
